---
title: "Breast Cancer Risk Prediction Using Bayesian Logistic Regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Breast cancer is one of the most common malignancies affecting women worldwide and remains a major public health concern due to its high incidence and risk of recurrence. Early detection and accurate prediction of cancer recurrence are crucial for improving patient prognosis and guiding clinical decision-making. As a result, statistical and machine learning approaches have been widely applied to analyze clinical and pathological data related to breast cancer outcomes.

This study aims to apply Bayesian Logistic Regression to analyze breast cancer recurrence using clinical attributes from the datase. Also leveraging the strengths of Bayesian inference in handling uncertainty and interpretability

# Data Preparation
```{r, message=FALSE, warning=FALSE}
#importing libraries used for the analysis

#install.packages(c("rjags", "coda", "dplyr", "pROC", "ggplot2", "bayesplot"))
library(rjags)
library(coda)
library(dplyr)
library(pROC)
library(ggplot2)
library(bayesplot)

```

The first step is to import our dataset

```{r, message=FALSE, warning=FALSE}
#importing dataset from .csv file
data_file <- "breast-cancer.csv"
data <- read.csv(data_file, header = TRUE)
head(data)

set.seed(123)
```

```{r, message=FALSE, warning=FALSE}
#setting features and target variable (M=Malignant, B=Benign) to binary form (1=M, 0=B)
data <- data %>%
  mutate(diagnosis_bin = ifelse(diagnosis == "M", 1, 0)) %>%
  select(diagnosis_bin, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean, concavity_mean, concave.points_mean, symmetry_mean, fractal_dimension_mean)
```

```{r, message=FALSE, warning=FALSE}
#separating feature (X) and target (Y)
Y <- data$diagnosis_bin
X <- data %>% select(-diagnosis_bin)
```

```{r, message=FALSE, warning=FALSE}
#scaling numerical features on 0-1 scale
X_scaled <- as.data.frame(scale(X))
N <- nrow(data)
K <- ncol(X_scaled)
```

# MCMC Modeling

```{r, message=FALSE, warning=FALSE}
#configuring MCMC Model
model_string <- "
model {
  # LIKELIHOOD (Bernoulli with Logit Link)
  for (i in 1:N) {
    Y[i] ~ dbern(p[i])
    logit(p[i]) <- alpha + inprod(beta[], X[i,])
  }

  # PRIOR DISTRIBUTION
  # Prior for Intercept (alpha)
  alpha ~ dnorm(0, 1)

  # Prior for Regression Coeff (beta)
  for (j in 1:K) {
    beta[j] ~ dnorm(0, 1) # Mean 0, Presisi 1
  }
}
"
```

```{r, message=FALSE, warning=FALSE}
#configuring data for JAGS
jags_data <- list(
  Y = Y,
  X = as.matrix(X_scaled),
  N = N,
  K = K
)
params <- c("alpha", "beta", "p")

inits <- list(
  list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 1),
  list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 2),
  list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 3)
)

```

```{r, message=FALSE, warning=FALSE}
#initializing JAGS model
model_jags <- jags.model(
  textConnection(model_string),
  data = jags_data,
  inits = inits,
  n.chains = 3,
)
```

```{r, message=FALSE, warning=FALSE}
#adding burn-in period
update(model_jags, 5000)
```

```{r, message=FALSE, warning=FALSE}
#taking samples from simulation
n_iter <- 15000
mcmc_samples <- coda.samples(
  model = model_jags,
  variable.names = params,
  n.iter = n_iter,
  thin = 10
)
```

# Posterior Analysis

```{r, message=FALSE, warning=FALSE}
# combining all chains for analysis
mcmc_combined <- as.mcmc(do.call(rbind, mcmc_samples))
```

```{r, message=FALSE, warning=FALSE}
# summarizing posterior statistics
summary_mcmc <- summary(mcmc_combined)
params <- c("alpha", paste0("beta[", 1:10, "]"))

# Filter the summary object to show only those rows
print(summary_mcmc$quantiles[params, c(2, 4, 5)])
```

```{r, fig.height=10}
# plotting posterior distribution
plot(mcmc_samples[, params])
```

# Model Evaluation

```{r, message=FALSE, warning=FALSE}
# evaluating using Posterior Predictive Checks (PPC)

# extracting probability samples
p_matrix <- as.matrix(mcmc_combined)
p_cols <- grep("^p", colnames(p_matrix))
p_samples <- p_matrix[, p_cols]

# generating Posterior Predictive Data (y_rep)
y_rep <- apply(p_samples, 2, function(p) rbinom(n = length(p), size = 1, prob = p))

# visualizing the Proportion of Malignancy
p1 <- ppc_stat(y = Y, yrep = y_rep, stat = "mean") + 
  labs(title = "PPC: Distribution of Predicted Malignancy Rates",
       x = "Proportion of Malignant Cases (mean Y)")

# showing the PPC Plot
print(p1)
```

```{r, message=FALSE, warning=FALSE}
# calculating the posterior mean for each patient's probability
posterior_mean_p <- colMeans(p_samples)

# converting into binary predictions (Y_pred = 1 if probability > 0.5)
Y_pred <- as.integer(posterior_mean_p > 0.5)

# evaluating using Confusion Matrix
conf_matrix <- table(Actual = Y, Predicted = Y_pred)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

print(conf_matrix)
cat("\n Model Accuracy: ", round(accuracy * 100, 2), "%\n")
```

```{r, message=FALSE, warning=FALSE}
# evaluating using AUC
roc_obj <- roc(Y, posterior_mean_p)
auc_value <- auc(roc_obj)
plot(roc_obj, main = paste("ROC Curve (AUC:", round(auc_value, 4), ")"))
grid()
```

# MCMC Diagnostics

```{r, message=FALSE, warning=FALSE}
# evaluating using Gelman–Rubin (PSRF)
mcmc_params_only <- mcmc_samples[, params]

gelman_results <- gelman.diag(mcmc_params_only, autoburnin = FALSE)
print(gelman_results)

# plotting Gelman Rubin
gelman.plot(mcmc_params_only, autoburnin = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# evaluating using Geweke Diagnostics

geweke_results <- geweke.diag(mcmc_params_only)
print(geweke_results)

# plotting Geweke
par(mfrow = c(2, 2))
geweke.plot(mcmc_params_only)
```

```{r, message=FALSE, warning=FALSE}
# calculating effective sample size
ess_vals <- effectiveSize(mcmc_params_only)
print(ess_vals)
```

